---
title: "Final Project"
author: "Preston Scibek"
date: "May 13, 2018"
output: html_document
---


## Intro
url_pitching = https://baseballsavant.mlb.com/statcast_search?hfPT=&hfAB=&hfBBT=&hfPR=&hfZ=&stadium=&hfBBL=&hfNewZones=&hfGT=R%7C&hfC=&hfSea=2018%7C&hfSit=&player_type=pitcher&hfOuts=&opponent=&pitcher_throws=&batter_stands=&hfSA=&game_date_gt=&game_date_lt=&team=BOS&position=&hfRO=&home_road=&hfFlag=&metric_1=&hfInn=&min_pitches=0&min_results=0&group_by=name&sort_col=pitches&player_event_sort=h_launch_speed&sort_order=desc&min_abs=0#results

url_batting = https://baseballsavant.mlb.com/statcast_search?hfPT=&hfAB=&hfBBT=&hfPR=&hfZ=&stadium=&hfBBL=&hfNewZones=&hfGT=R%7C&hfC=&hfSea=2018%7C&hfSit=&player_type=batter&hfOuts=&opponent=&pitcher_throws=&batter_stands=&hfSA=&game_date_gt=&game_date_lt=&team=BOS&position=&hfRO=&home_road=&hfFlag=&metric_1=&hfInn=&min_pitches=0&min_results=0&group_by=name&sort_col=pitches&player_event_sort=h_launch_speed&sort_order=desc&min_abs=0#results

This data comes from the mlb statcast dataset. The data can be found at the above urls by clicking on the icon above the graphs column in the table. Then load the data as shown below.

```{r setup, include=TRUE,  message=F, warning=F}
library(tidyverse)
library(stringr)
library(rvest)
library(ggplot2)
library(plotly)
library(tree)
library(randomForest)
library(png)
library(caret) 
library(rpart)

final_pitching_data_table <- read_csv("redsox_pitching_5132018.csv")
final_batting_data_table <- read_csv("redsox_batting_5132018.csv")

head(final_pitching_data_table)

head(final_batting_data_table)
```


## Create ID Tables
If we look at the pitching and batting data we will notice that there are fields with ids that are not translated to names. So we will use the batting and pitching data to generate ids for each person. Since the player_name is the name of the batter/pitcher in their respective data sets we should be able to get all the ids by extracting the player_name with the batter field for the batting table and the player_name field with the pitcher field for the pitching data.

```{r get id table, include=TRUE, message=F, warning=F}
batter_id_table <- final_batting_data_table %>%
  group_by(player_name, batter) %>%
  select(player_name, batter) %>%
  distinct

pitcher_id_table <- final_pitching_data_table %>%
  group_by(player_name, pitcher) %>%
  select(player_name, pitcher) %>%
  distinct

head(batter_id_table)

head(pitcher_id_table)
```

## Get Mookie Betts Data
Mookie Betts is having a standout season. I would like to focus on data regarding Mookie's at bats.

```{r get mookie data,  message=F, warning=F}
mookie_data <- final_batting_data_table %>%
  filter(player_name=="Mookie Betts")

head(mookie_data)

```

## Events
```{r mookie at bat events,  message=F, warning=F}
mookie_data <- mookie_data %>%
  mutate(rbi=post_bat_score-bat_score)
  
mookie_data %>%
  group_by(events) %>%
  summarize(count=n())
```
Lets look at the at bats but grouped based on pitch type.
```{r at bat pitch type,  message=F, warning=F}
mookie_data %>%
  group_by(events, pitch_name) %>%
  summarize(count=n())
```

Some raw data
links to column names https://fastballs.wordpress.com/category/pitchfx-glossary/

pfx_x: the horizontal movement, in inches, of the pitch between the release point and home plate, as compared to a theoretical pitch thrown at the same speed with no spin-induced movement. This parameter is measured at y=40 feet regardless of the y0 value.

pfx_z: the vertical movement, in inches, of the pitch between the release point and home plate, as compared to a theoretical pitch thrown at the same speed with no spin-induced movement. This parameter is measured at y=40 feet regardless of the y0 value.

plate_x: the left/right distance, in feet, of the pitch from the middle of the plate as it crossed home plate. The PITCHf/x coordinate system is oriented to the catcher’s/umpire’s perspective, with distances to the right being positive and to the left being negative.

plate_z: the height of the pitch in feet as it crossed the front of home plate.


```{r at bats,  message=F, warning=F}
mookie_bat_data <- mookie_data %>%
  select(events, pitch_name, release_speed, pitcher, bb_type, pfx_x, pfx_z, plate_x, plate_z, launch_angle, launch_speed, hit_distance_sc, sz_top, sz_bot)
mookie_bat_data %>%
  filter(events=="home_run") %>%
  head()
```

Before we can analyze and graph this we need to clean it up.
```{R cleaning,  message=F, warning=F}
mookie_bat_data <- mookie_bat_data %>%
  type_convert(cols(release_speed=col_double(), pfx_x=col_double(), pfx_z=col_double(), plate_x=col_double(), plate_z=col_double(), launch_angle=col_double(), launch_speed=col_double(), hit_distance_sc=col_integer(), sz_top=col_double(), sz_bot=col_double()))

mookie_bat_data$pitch_name <- as.factor(mookie_bat_data$pitch_name)
mookie_bat_data$events <- as.factor(mookie_bat_data$events)
mookie_bat_data$bb_type <- as.factor(mookie_bat_data$bb_type)
head(mookie_bat_data)
```
```{r graphing,  message=F, warning=F}
strike_zone_top <- mean(mookie_bat_data$sz_top, na.rm=TRUE)
strike_zone_bot <- mean(mookie_bat_data$sz_bot, na.rm=TRUE)
strike_zone_front <- 17/2.0/12


graphed <- ggplot() + geom_rect(aes(xmin=-strike_zone_front, xmax=strike_zone_front, ymin=strike_zone_bot, ymax=strike_zone_top), color="black", fill=NA) + geom_point(data=mookie_bat_data, aes(x=plate_x, y=plate_z, text=sprintf("events: %s<br>Pitch Name: %s<br>Release Speed: %s<br>Launch Angle: %s", events, pitch_name, release_speed, launch_angle))) +
  scale_x_continuous(name="Catchers View | left/right distance in feet of center of plate", limits=c(-4, 4), breaks=seq(from=-4, to=4, by=2)) + scale_y_continuous(name="Height above plate In Feet", limits=c(-2, 6), breaks=seq(from=-2, to=6, by=1))

ggplotly(graphed) 
```
This is rather ugly. Lets do pitches that result in action and color code it by the event that happens.
```{r hits,  message=F, warning=F}
action <- mookie_bat_data %>%
  filter(events!="null")

graphed <- ggplot() + geom_rect(aes(xmin=-strike_zone_front, xmax=strike_zone_front, ymin=strike_zone_bot, ymax=strike_zone_top), color="black", fill=NA) + geom_point(data=action, aes(x=plate_x, y=plate_z, text=sprintf("events: %s<br>Pitch Name: %s<br>Release Speed: %s<br>Launch Angle: %s", events, pitch_name, release_speed, launch_angle), color=events)) +
  scale_x_continuous(name="Catchers View | left/right distance in feet of center of plate", limits=c(-4, 4), breaks=seq(from=-4, to=4, by=2)) + scale_y_continuous(name="Height above plate In Feet", limits=c(-2, 6), breaks=seq(from=-2, to=6, by=1))

ggplotly(graphed) 
```
Mookie leads the league in HRS. Let's filter by home runs and then graph the hrs colored by pitch.
```{r hrs,  message=F, warning=F}
action <- mookie_bat_data %>%
  filter(events == "home_run")

graphed <- ggplot() + geom_rect(aes(xmin=-strike_zone_front, xmax=strike_zone_front, ymin=strike_zone_bot, ymax=strike_zone_top), color="black", fill=NA) + geom_point(data=action, aes(x=plate_x, y=plate_z, text=sprintf("events: %s<br>Pitch Name: %s<br>Release Speed: %s<br>Launch Angle: %s", events, pitch_name, release_speed, launch_angle), color=pitch_name)) +
  scale_x_continuous(name="Catchers View | left/right distance in feet of center of plate", limits=c(-4, 4), breaks=seq(from=-4, to=4, by=2)) + scale_y_continuous(name="Height above plate In Feet", limits=c(-2, 6), breaks=seq(from=-2, to=6, by=1))

ggplotly(graphed) 
```

Lets try to predict the outcome of a pitch.
```{r predict,  message=F, warning=F}
set.seed(1234)
mookie_bat_data_converted <- mookie_bat_data %>%
  filter(events!="null") %>%
  filter(events!="intent_walk") %>%
  filter(!is.na(plate_x)) %>%
  select(events, pitch_name, release_speed, plate_x, plate_z)




mookie_bat_data_converted 
mookie_tree <- tree(events~pitch_name+plate_x+plate_z+release_speed, data=mookie_bat_data_converted)

# We will save the plot as a png to deal with overlapping graphics
png("image.png", width = 1200, height = 700)
plot(mookie_tree)
text(mookie_tree, pretty=1, cex=1.1)
dev.off()

img <- readPNG('image.png')
grid::grid.raster(img)
```
This tree shows us the what the likelihood outcome of a pitch is based on location, type, and speed.

Lets try and use a Random Forest
```{r rf,  message=F, warning=F}
mookie_bat_data_converted$pitch_name = as.numeric(mookie_bat_data_converted$pitch_name)

mookie_bat_data_converted <- droplevels(mookie_bat_data_converted)

train_set <- mookie_bat_data_converted %>% 
  group_by(events) %>%
  sample_frac(.6) %>%
  ungroup()

test_set <- mookie_bat_data_converted %>%
  anti_join(train_set)

train_set
test_set
mookie_rf <- randomForest(events~pitch_name+plate_x+plate_z, data=train_set)


test_predictions <- predict(mookie_rf, newdata=test_set)
cm <- table(pred=test_predictions, observed=test_set$events)

confusionMatrix(cm)
```
46% accurate is really bad. So there might not be enough data on mookie to predict properly.
What if we try on all the red sox batters with as many predictors as we can get. We forgot that a 3-0 count is never gonna result in a strikeout whereas a 0-2 will never get a walk, and if no one is one base there can't be a doulbe play. So we add as many predictors as we can.       
```{r Type converting,  message=F, warning=F}
# We filter out na's and select our columns.
final_batting_data_table_converted <- final_batting_data_table %>%
  mutate(release_speed=replace(release_speed, release_speed=="null", 0))%>%
   mutate(pfx_x=replace(pfx_x, pfx_x=="null", 0))%>%
   mutate(pfx_z=replace(pfx_z, pfx_z=="null", 0))%>%
   mutate(plate_x=replace(plate_x, plate_x=="null", 0))%>%
   mutate(plate_z=replace(plate_z, plate_z=="null", 0))%>%
   mutate(on_1b=replace(on_1b, on_1b=="null", 0))%>%
   mutate(on_2b=replace(on_2b, on_2b=="null", 0))%>%
   mutate(on_3b=replace(on_3b, on_3b=="null", 0))%>%
   mutate(hit_location=replace(hit_location, hit_location=="null", 0))%>%
  mutate(hit_distance_sc=replace(hit_distance_sc, hit_distance_sc=="null", 0))%>%
  mutate(launch_angle=replace(launch_angle, launch_angle=="null", 0))%>%
  mutate(launch_speed=replace(launch_speed, launch_speed=="null", 0))%>%
  
  type_convert(cols(release_speed=col_double(), pfx_x=col_double(), pfx_z=col_double(), plate_x=col_double(), plate_z=col_double(), on_1b=col_double(), on_2b=col_double(), on_3b=col_double(), hit_distance_sc=col_double(), launch_angle=col_double(), launch_speed=col_double(), hit_location=col_double())) %>%
  filter(events!="intent_walk") %>%
  filter(events!="null") %>%
  filter(!is.na(plate_x)) %>%
  filter(!is.na(plate_z)) %>%
  filter(!is.na(pitch_name)) %>%
  filter(!is.na(release_speed)) %>%
  mutate(rbi=post_bat_score-bat_score) %>%
  select(events, pitch_name, release_speed, plate_x, plate_z, pfx_x, pfx_z, batter, pitcher, balls, strikes, outs_when_up, on_1b, on_2b, on_3b, hit_location, pfx_x, pfx_z, hit_distance_sc, launch_speed, launch_angle, pitch_number, rbi)
```

```{r prediction on all}
final_batting_data_table_converted$pitch_name <- as.factor(final_batting_data_table_converted$pitch_name)
final_batting_data_table_converted$events <- as.factor(final_batting_data_table_converted$events)

final_batting_data_table_converted$on_1b[is.na(final_batting_data_table_converted$on_1b)] <- 0
final_batting_data_table_converted$on_2b[is.na(final_batting_data_table_converted$on_2b)] <- 0
final_batting_data_table_converted$on_3b[is.na(final_batting_data_table_converted$on_3b)] <- 0

# We convert the pitch name to numeric
final_batting_data_table_converted$pitch_name = as.numeric(final_batting_data_table_converted$pitch_name)

# We drop unused levels
final_batting_data_table_converted <- droplevels(final_batting_data_table_converted)

# Use 60/40 split for train data / test data
train_set <- final_batting_data_table_converted %>% 
  group_by(events) %>%
  sample_frac(.8) %>%
  ungroup()

test_set <- final_batting_data_table_converted %>%
  anti_join(train_set)

train_set
test_set
final_rf <- randomForest(events~batter+pitcher+pitch_name+plate_x+plate_z+release_speed+balls+strikes+outs_when_up+on_1b+on_2b+on_3b+hit_location+pfx_x+pfx_z+hit_distance_sc+launch_speed+launch_angle+pitch_number+rbi, data=train_set)

test_predictions <- predict(final_rf, newdata=test_set)
cm <- table(pred=test_predictions, observed=test_set$events)

confusionMatrix(cm)

```
85% accuracy is a lot better. By adding more predictors we were able to drastically improve our accuracy.
